---
title: "smic_initialQC"
output: html_document
date: "2026-03-12"
---

First we need to load the necessary files to create our phyloseq object
***Note that singletons were removed previously so are already taken out in these files.
**We also removed swabs from sites and species not used in this paper from the metadata, as there were additional samples included in these sequencing runs that we did not end up focusing on here.

```{r load files}

setwd("C:/Users/Carolyn/Documents/seastar-microbiome-paper/data")

asvfile <-read.csv("asvtable_NCM_smiconly-runs1to8.csv", row.names=1, header = TRUE) 
taxfile <-as.matrix(read.csv("smic_runs1to8_silva138-1_taxonomy.csv",row.names=1,header = TRUE)) 
mapfile <-read.csv("metadata_updated_2026.csv",row.names = 1, header=TRUE)

mapfile$Date <- as.Date(mapfile$Date, format="%Y-%m-%d")

```

Now we combine these 3 files to create a phyloseq object

```{r phyloseq object}

library(phyloseq)
library(data.table)
library(ggplot2)

OTU = otu_table(asvfile, taxa_are_rows = TRUE)
TAX = tax_table(taxfile)
MAP = sample_data(mapfile)

sswd16S <- merge_phyloseq(OTU, TAX, MAP)

sswd16S_reads <- data.table(as(sample_data(sswd16S), "data.frame"),
                 TotalReads = sample_sums(sswd16S), keep.rownames = TRUE)

min(sswd16S_reads$TotalReads) #2
mean(sswd16S_reads$TotalReads) #38405.8
median(sswd16S_reads$TotalReads) #30486
max(sswd16S_reads$TotalReads) #995940

sswd16S
#13321 taxa and 1631 samples

#plot the reads across samplesm with a y-intercept line of 150,000 (potential upper read limit cut-off)
ggplot(sswd16S_reads, aes(x = rownames(sswd16S_reads), y = TotalReads, color = Host)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_hline(yintercept = 150000, linetype = "solid") +
  xlab("Samples") +
  ylab("Counts") +
  ggtitle("All data runs 1-8: Read counts per sample, line at 150,000 reads")

```


Now we need to do some QC on the dataset before doing any visualizations or analyses.

Those ASVs that have an 'unknown' or 'bacteria' only annotation as we cannot confirm if they are truly bacteria or unannotated mitochondria or chloroplasts. Also prudent to remove Proteobacteria only and Alphaproteobacteria only ASV because we cannot confirm or deny if they are mitochondria that just are not annotated.

```{r remove unknown annnotations}

library(dplyr)
library(microViz)

sswd16S #13321 taxa and 1631 samples

#here we are propagating the last assigned taxonomic level down to species
sswd_fix<- sswd16S %>%
 tax_fix(
  min_length = 4,
  unknowns = c("", "NA","unknown","uncultured","unassigned","endosymbionts", "endosymbiont_of", "endosymbionts Genus", "Unknown_Family"),
  sep = " ", anon_unique = TRUE,
  suffix_rank = "classified")


#get rid of uninformative groups
sswd1<-sswd_fix %>% tax_select(tax_list = "Unassigned Domain", deselect = TRUE) %>% 
  tax_select(tax_list = "Bacteria Domain", deselect = TRUE) 

sswd1   #12951 taxa and 1631 samples

#we also want to get rid of Proteobacteria Phylum (class) and Alphaproteobacteria Class (family) as these unknown reads could be mitochondrial. I wonder if we should also get rid of Cyanobacteriia Class as it could be unassigned chloroplasts? Think about this.

sswd2<-sswd1 %>% tax_select(tax_list = "Proteobacteria Phylum", deselect = TRUE) %>% tax_select(tax_list = "Alphaproteobacteria Class", deselect = TRUE)

sswd2 #12758 taxa and 1631 samples

```

Double checking we have the sites and sample types of interest in the dataset (data were filtered as sequencing runs contained some other species and sites)

```{r}

unique(get_variable(sswd2, "Site"))
#annex, shellaligan pass, moulds bay, north beach ri, west beach ri, fifth beach -- only intertidal sites looks good     #N/A is negative controls

unique(get_variable(sswd2, "Host"))
#"Pisaster ochraceus"     "Dermasterias imbricata" "Seawater" 

#see how many true samples and negatives we have remaining in this dataset:
table(sample_data(sswd2)$SampleType2)

#SeawaterExtNeg SeawaterPCRNeg SeawaterSample     SwabExtNeg     SwabPCRNeg     SwabSample 
    #     11              1             74             77             15           1453 


```

It makes sense to run decontam on the seawater and swab samples SEPARATELY since the extraction methods were quite different and they have their own associated extraction and PCR negatives. If we run them together, ASVs abundant in seawater but absent in low-biomass swabs may be falsely flagged as contaminants and ASVs abundant in swabs but low in seawater may be ignored as contaminants. So we need to subset the data accordingly and then re-merge after decontam.

```{r}
sswd_seawater <- ps_filter(sswd2, SampleType2 %in% c("SeawaterSample", "SeawaterExtNeg", "SeawaterPCRNeg"))
sswd_seawater #3126 taxa and 86 samples 

sswd_stars <- ps_filter(sswd2, SampleType2 %in% c("SwabSample", "SwabExtNeg", "SwabPCRNeg"))
sswd_stars #11270 taxa and 1545 samples

```


Now for the decontam package which is used to identify contaminant ASVs based on our negative controls. There are two thresholds (0.1 and 0.5) and here we are using the more stringent 0.5 cut-off largely because most samples were extracted with plate kits which are more subject to cross contamination. Since different extraction methods were used for seawater and swabs and they represent quite different sample types with potentially different microbial communities, we will run decontam separately for each and then merge the decontaminated phyloseq objects

https://benjjneb.github.io/decontam/vignettes/decontam_intro.html

```{r decontam - sea star swabs}

#first running decontam on sea star swabs and then will do seawater separately and merge the two cleaned phyloseqs
library(ggplot2)
library(decontam); packageVersion("decontam")
library(microViz)

#using decontam package to control for contaminants identified in negative controls

sample_data(sswd_stars)$is.neg <- sample_data(sswd_stars)$SampleNeg == "Negative" 
#negative here includes both extraction and pcr negs from SWAB samples only in this case

#here we are using a more aggressive threshold of 0.5 rather than the default of 0.1
#this will identify as contaminants all sequences that are more prevalent in negative controls than in positive sample

contamdf.prev05_stars <- isContaminant(sswd_stars, method="prevalence", neg="is.neg", threshold=0.5)

table(contamdf.prev05_stars$contaminant)
#FALSE  TRUE 
#10926   344    

# Make phyloseq object of presence-absence in negative controls and true sea star samples
ps.pa_stars <- transform_sample_counts(sswd_stars, function(abund) 1*(abund>0))
ps.pa_stars.neg <- prune_samples(sample_data(ps.pa_stars)$SampleNeg == "Negative", ps.pa_stars)
ps.pa_stars.pos <- prune_samples(sample_data(ps.pa_stars)$SampleNeg == "Sample", ps.pa_stars)


# Make data.frame of prevalence in positive and negative samples using 0.5 threshold
df.pa0.5_stars <- data.frame(pa.pos=taxa_sums(ps.pa_stars.pos), pa.neg=taxa_sums(ps.pa_stars.neg),
                    contaminant=contamdf.prev05_stars$contaminant)

ggplot(data=df.pa0.5_stars, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)") + theme_light() + scale_color_brewer(palette = "Dark2") + labs(title="0.5 More Aggresive Threshold")


#creating a new decontaminated phyloseq object for the swab samples

decontam_swabs <- prune_taxa(!contamdf.prev05_stars$contaminant, sswd_stars)

decontam_swabs #<-<-phyloseq object with contamination removed using the higher threshold PREVALENCE method
# 10926 taxa and 1545 samples

```

Now same thing for seawater 
```{r decontam - seawater}

#first running decontam on sea star swabs and then will do seawater separately and merge the two cleaned phyloseqs
library(ggplot2)
library(decontam); packageVersion("decontam")
library(microViz)

#using decontam package to control for contaminants identified in negative controls

sswd_seawater
sample_data(sswd_seawater)$is.neg <- sample_data(sswd_seawater)$SampleNeg == "Negative" 
#negative here includes both extraction and pcr negs from Seawater samples only in this case

#here we are using a more aggressive threshold of 0.5 rather than the default of 0.1
#this will identify as contaminants all sequences that are more prevalent in negative controls than in positive sample

contamdf.prev05_h20 <- isContaminant(sswd_seawater, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf.prev05_h20$contaminant)
#FALSE  TRUE 
# 3096    30 

# Make phyloseq object of presence-absence in negative controls and true sea star samples
ps.pa_h20 <- transform_sample_counts(sswd_seawater, function(abund) 1*(abund>0))
ps.pa_h20.neg <- prune_samples(sample_data(ps.pa_h20)$SampleNeg == "Negative", ps.pa_h20)
ps.pa_h20.pos <- prune_samples(sample_data(ps.pa_h20)$SampleNeg == "Sample", ps.pa_h20)


# Make data.frame of prevalence in positive and negative samples using 0.5 threshold
df.pa0.5_seawater <- data.frame(pa.pos=taxa_sums(ps.pa_h20.pos), pa.neg=taxa_sums(ps.pa_h20.neg),
                    contaminant=contamdf.prev05_h20$contaminant)

ggplot(data=df.pa0.5_seawater, aes(x=pa.neg, y=pa.pos, color=contaminant)) + geom_point() +
  xlab("Prevalence (Negative Controls)") + ylab("Prevalence (True Samples)") + theme_light() + scale_color_brewer(palette = "Dark2") + labs(title="0.5 More Aggresive Threshold")


#creating a new decontaminated phyloseq object for the seawater h20 samples

decontam_seawater <- prune_taxa(!contamdf.prev05_h20$contaminant, sswd_seawater)

decontam_seawater #<-<-phyloseq object with contamination removed using the higher threshold PREVALENCE method
#3096 taxa and 86 samples

```

now we want to merge these phyloseq objects back together

```{r merge swab and seawater decontam}

decontam_merged <- merge_phyloseq(decontam_seawater, decontam_swabs)
decontam_merged
table(sample_data(decontam_merged)$Host)

```

Now do a final filtering of low read-count samples (i.e. similar to negatives), and remove any remaining negative controls

```{r}

decontam_filter <- prune_samples(sample_sums(decontam_merged) >= 7000, decontam_merged)
decontam_filter

decontam_filter_nonegs <- ps_filter(decontam_filter, SampleNeg != "Negative")
decontam_filter_nonegs  #11452 taxa and 1364 samples

#saving this phyloseq object
saveRDS(decontam_filter_nonegs, file = "C:/Users/Carolyn/Documents/intertidal-seastar-swabs/code/analysis chunks/decontam_filter_nonegs.rds")

```

For some analyses, e.g. beta diversity, indval, we will want to also filter the data to remove rare ASVs. I will save this as a second RDS file

```{r}

#for beta diversity we want to filter out ASVs present in less than 5% of samples

# Number of samples
nsamples_total <- nsamples(decontam_filter_nonegs)

# Calculate prevalence (number of samples where ASV > 0)
prev_counts <- apply(otu_table(decontam_filter_nonegs), 1, function(x) sum(x > 0))

# Convert to proportion
prev_prop <- prev_counts / nsamples_total

#Setting prevelance level
decontam_filter_nonegs_prev <- prune_taxa(prev_prop >= 0.01, decontam_filter_nonegs)

reads_post_prev <- data.table(as(sample_data(decontam_filter_nonegs_prev), "data.frame"),
                 TotalReads = sample_sums(decontam_filter_nonegs_prev), keep.rownames = TRUE)

decontam_filter_nonegs_prev  #3839 taxa and 1364 samples

ggplot(reads_post_prev , aes(x = rownames(reads_post_prev), y = TotalReads, color = Host)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Samples") +
  ylab("Counts") +
  ggtitle("Filtered data: Read counts per sample")

#saving this phyloseq object

saveRDS(decontam_filter_nonegs_prev, file = "C:/Users/Carolyn/Documents/intertidal-seastar-swabs/code/analysis chunks/decontam_filter_nonegs_prev.rds")

```


