---
title: "A bit of indicator species analysis"
author: "Colleen Kellogg"
date: "`r Sys.Date()`"
output: html_document
---

```{r message = FALSE}
####################### libraries #######################
library(phyloseq)
library(tidyverse)
library(ggplot2)
library(vegan)
library(data.table)
library(indicspecies)
library(colorspace)
library(dplyr)
library(tidyr)
library(microViz)

```

### Indicator Species Analysis

Calculate Host Indicators

Your standardized/normalized (whatever your preference for working with compositional data) ASV table needs to have samples as columns and ASVs as rows for multipatt. 

```{R Load and prepare data}

smic_indicator <- readRDS("C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/decontam_filter_nonegs_prev.rds")


#need to rename taxa because otherwise R will add an X to the start of the MD5sum versions of the ASV names
# get taxa sums and sort
smic_indicator_sorted <- smicpf %>% 
  tax_sort(by = sum)
  
taxa_names(smic_indicator_sorted) <- paste0("ASV", seq(ntaxa(smic_indicator_sorted)))

smic_healthyseawater<-smic_indicator_sorted %>% subset_samples(WastingScore == "0" | Host == "Seawater") #subset to keep only healthy stars and seawater

#make sure we get everything:
# sample number summary
table(sample_data(smic_healthyseawater)$Host)

# Dermasterias imbricata     Pisaster ochraceus               Seawater 
#                    416                    530                     72 

##### ASV level inputs #####

smic_asvs <- data.frame((otu_table(smic_healthyseawater))) #extract asv table from phyloseq object
smic_meta <- data.frame(sample_data(smic_healthyseawater)) %>% select(Host,SiteType,SiteID,WastingScore,WastingCategory,Year,Season,YearMonth)

#Normalize
smicR <- decostand(smic_asvs, method = "total", MARGIN = 1)

#check to make sure that standardization worked
rowSums(smicR) #should all be 1


#### Genus level ####

smic_gen <-speedyseq::tax_glom(smicpf, taxrank = "Genus")

smic_gen_sorted <- smic_gen %>% 
  tax_sort(by = sum)

taxa_names(smic_gen_sorted) <- paste0("Genus", seq(ntaxa(smic_gen_sorted)))

smic_healthyseawater_gen<-smic_gen_sorted %>% subset_samples(WastingScore == "0" | Host == "Seawater") #subset to keep only healthy stars and seawater

#make sure we get everything:
# sample number summary
table(sample_data(smic_healthyseawater_gen)$Host)

smic_genera <- data.frame((otu_table(smic_healthyseawater_gen))) #extract asv table from phyloseq object

#Normalize
smic_genR <- decostand(smic_genera, method = "total", MARGIN = 1)

#check to make sure that standardization worked
rowSums(smic_genR) #should all be 1


#### Indicator Groupings ####

#create a grouping variables
hostgroups <- structure(c(smic_meta$Host), .Names = c(row.names(smic_meta)))

```


```{r Genus level Indicator Species Analysis}

#### run indicator species analysis ####
host.indval.gen = multipatt(smic_genR, hostgroups, control = how(nperm=999), duleg = FALSE) #set duleg to FALSE if you want to also allow for two-group indicators (e.g. derm + pisaster)

summary(host.indval.gen)
attributes(host.indval.gen)

isa.genera<-host.indval.gen$sign

#write.csv(isa.genera, "SMIC-multipatt_genera_indval_results.csv") #save the results

#### filter data frame for significant indicators ####
#In order to be included we want to retain only indicators with a p value of 0.005 or less and a stat of 0.7 or higher.

isa_tbl_genera<-tibble(rownames_to_column(isa.genera, var = "ASVID")) %>% rename(c("Seawater" = "s.Seawater", "Dermasterias" = "s.Dermasterias imbricata", "Pisaster" = "s.Pisaster ochraceus")) %>% mutate("ind_group"=index) %>%  mutate(ind_group = replace(ind_group, ind_group == 1, "Dermasterias")) %>% mutate(ind_group = replace(ind_group, ind_group == 2, "Pisaster")) %>% mutate(ind_group = replace(ind_group, ind_group == 3, "Seawater"))

#how many signficant indicator ASVs are there in your different sample groups (you can change the parameters for signficance)
sum(isa_tbl_genera$index == 3 & isa_tbl_genera$stat >= 0.7 & isa_tbl_genera$p.value <= 0.001) 
#[1] 67 = seawater indicators

sum(isa_tbl_genera$index == 2 & isa_tbl_genera$stat >= 0.7 & isa_tbl_genera$p.value <= 0.001)
#[1] 8 = pisaster indicators

sum(isa_tbl_genera$index == 1 & isa_tbl_genera$stat >= 0.7 & isa_tbl_genera$p.value <= 0.001)
#[1] 4 = derm indicators

#make a new column denoting indicator status

isa_tbl_genera <- isa_tbl_genera %>% mutate("isInd" = case_when(
  stat >= 0.7 & p.value <= 0.001 ~ "Yes",
  stat < 0.7 | p.value > 0.001 ~ "No"
  ))

isa_tbl_genera <-isa_tbl_genera %>% mutate("Ind_status" = ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Seawater", 'Seawater', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Pisaster", 'Pisaster', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Dermasterias", 'Dermasterias', 'No'))))

table(isa_tbl_genera$isInd)

isa_tbl_genera

#### merge indicator status with taxonomy table ####
#now we need to add a column about indicator status to the taxonomy table

Ind_tbl_genera <- isa_tbl_genera %>% select(ASVID, Ind_status)

### map back to the full data frame, including diseased stars ###

smic_gen_taxa<-data.frame(tax_table(smic_gen_sorted)) %>% rownames_to_column() %>% dplyr::rename("ASVID" = "rowname") 

smic_genera_tax_ind<-left_join(smic_gen_taxa, Ind_tbl_genera)
smic_genera_tax_ind

smic_genera_tax_ind<-as.matrix(column_to_rownames(as.data.frame(smic_genera_tax_ind), var = "ASVID"))

#### update taxonomy table in phyloseq object ####
smic_isa.genus <- smic_gen_sorted #rename original phyloseq object
tax_table(smic_isa.genus) <- smic_genera_tax_ind #update taxa table to include ISA

```


#### RE RUNNING INDICATOR ANALYIS ALLOWING OVERLAPPING INDICATORS AMONG GROUPS ####

```{r}
#### run indicator species analysis ####
host.indval.gen.v2 = multipatt(smic_genR, hostgroups, control = how(nperm=999), duleg = FALSE) #set duleg to FALSE if you want to also allow for two-group indicators (e.g. derm + pisaster)

summary(host.indval.gen.v2)
attributes(host.indval.gen.v2)

isa.genera.v2<-host.indval.gen.v2$sign

#write.csv(isa.genera.v2, "SMIC-multipatt_genera_indval_results.csv") #save the results

#### filter data frame for significant indicators ####
#In order to be included we want to retain only indicators with a p value of 0.005 or less and a stat of 0.7 or higher.

isa_tbl_genera.v2<-tibble(rownames_to_column(isa.genera.v2, var = "ASVID")) %>% rename(c("Seawater" = "s.Seawater", "Dermasterias" = "s.Dermasterias imbricata", "Pisaster" = "s.Pisaster ochraceus")) %>% mutate("ind_group"=index) %>%  mutate(ind_group = replace(ind_group, ind_group == 1, "Dermasterias")) %>% mutate(ind_group = replace(ind_group, ind_group == 2, "Pisaster")) %>% mutate(ind_group = replace(ind_group, ind_group == 3, "Seawater")) %>%
mutate(ind_group = replace(ind_group, ind_group == 4, "Dermasterias+Pisaster")) %>%
mutate(ind_group = replace(ind_group, ind_group == 5, "Dermasterias+Seawater")) %>%
mutate(ind_group = replace(ind_group, ind_group == 6, "Pisaster+Seawater")) %>%  
mutate(ind_group = replace(ind_group, ind_group == 7, "All"))

#how many signficant indicator ASVs are there in your different sample groups (you can change the parameters for signficance)
sum(isa_tbl_genera.v2$index == 3 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001) 
#[1] 67 = seawater indicators

sum(isa_tbl_genera.v2$index == 2 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] 8= pisaster indicators

sum(isa_tbl_genera.v2$index == 1 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] 4 = derm indicators

sum(isa_tbl_genera.v2$index == 4 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] 7 = derm+pis indicators

sum(isa_tbl_genera.v2$index == 5 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] 0 = derm+sw indicators

sum(isa_tbl_genera.v2$index == 6 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] 1 = pis+sw indicators

sum(isa_tbl_genera.v2$index == 7 & isa_tbl_genera.v2$stat >= 0.7 & isa_tbl_genera.v2$p.value <= 0.001)
#[1] NA = pis+derm+sw indicators. I'm not sure why but the ALL (group 7) indicators do not have p-values.



#make a new column denoting indicator status

isa_tbl_genera.v2 <- isa_tbl_genera.v2 %>% mutate("isInd" = case_when(
  stat >= 0.7 & p.value <= 0.001 ~ "Yes",
  stat < 0.7 | p.value > 0.001 ~ "No"
  ))

isa_tbl_genera.v2 <-isa_tbl_genera.v2 %>% mutate("Ind_status" = ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Seawater", 'Seawater', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Pisaster", 'Pisaster', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Dermasterias", 'Dermasterias', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Dermasterias+Seawater", 'Dermasterias+Seawater', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Dermasterias+Pisaster", 'Dermasterias+Pisaster',ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Pisaster+Seawater", 'Pisaster+Seawater', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "All", 'All', 'No'))))))))

table(isa_tbl_genera.v2$isInd)

isa_tbl_genera.v2 #all the NAs are the index=7 / All

#### merge indicator status with taxonomy table ####
#now we need to add a column about indicator status to the taxonomy table

Ind_tbl_genera.v2 <- isa_tbl_genera.v2 %>% select(ASVID, Ind_status)

### map back to the full data frame, including diseased stars ###

smic_gen_taxa.v2<-data.frame(tax_table(smic_gen_sorted)) %>% rownames_to_column() %>% dplyr::rename("ASVID" = "rowname") 

smic_genera_tax_ind.v2<-left_join(smic_gen_taxa.v2, Ind_tbl_genera.v2)
smic_genera_tax_ind.v2

smic_genera_tax_ind.v2<-as.matrix(column_to_rownames(as.data.frame(smic_genera_tax_ind.v2), var = "ASVID"))

#### update taxonomy table in phyloseq object ####
smic_isa.genus.v2 <- smic_gen_sorted #rename original phyloseq object
tax_table(smic_isa.genus.v2) <- smic_genera_tax_ind.v2 #update taxa table to include IS


```



Filter and plot the indicator species results. 

I'm including this because it shows how I mapped if any ASV was an indicator back to the phyloseq object, which can be handy for plotting and using packages that leverage the phyloseq data structure (ampVis2, microViz, ...)
```{r a few quick plots }

#input = isa; indicator species results from chunk above

# smic_isa.asv #indicator ASVs
# smic_isa.genus #indicator genera



#now you can filter by indicator and just plot the composition of indicator taxa, etc.
#smic_isa.asvm<- speedyseq::psmelt(smic_isa.asv %>%  
  #transform_sample_counts(function(x) {x/sum(x)} ))

smic_isa.genusm<- speedyseq::psmelt(smic_isa.genus.v2 %>%  
  transform_sample_counts(function(x) {x/sum(x)} ))

#### quick bubble plot for indicator genera ####

smic_gen.ind_ave <- tibble(smic_isa.genusm) %>% group_by(OTU,Host, Phylum, Class, Order, Family, Genus, Ind_status, Year, YearMonth, Season, SiteType, WastingScore, WastingCategory, DermisType2) %>% summarise(ave_abun = mean(Abundance)) 
#tidyr:: if unite doesn't work

smic_gen.ind_ave
#write.csv(smic_gen.ind_ave, "SMIC-multipatt_genera_indval_results.csv")

smic_gen.ind_ave <-  smic_gen.ind_ave %>% 
  mutate(Year_Season = paste(Year, Season, sep = "_")) %>% 
  mutate(Year_Season = factor(Year_Season, levels = c("2020_Winter", "2021_Winter", "2022_Winter", "2023_Winter", 
"2020_Spring", "2021_Spring", "2022_Spring", "2023_Spring", "2020_Summer", "2021_Summer", "2022_Summer", "2023_Summer", "2020_Fall", "2021_Fall", "2022_Fall", "2023_Fall"))) %>% 
  mutate(Season = factor(Season, levels = c("Winter", "Spring", "Summer", "Fall")))


ind_genera_plot_season<-smic_gen.ind_ave  %>%
  tidyr::unite(GENIND, Genus, Ind_status, sep = "_", remove = FALSE) %>%
  filter(Host != "Seawater") %>%
  filter(ave_abun > 0) %>%
  filter(WastingScore == "0") %>%
  filter(Ind_status %in% c("Dermasterias+Pisaster", "Dermasterias", "Pisaster")) %>%
ggplot(aes(x = Season, y = GENIND, size = ave_abun, color = Class)) +
  geom_point(alpha = 0.7) +
  scale_color_discrete_sequential(palette = "Viridis") +
  scale_size_continuous(range = c(0.5, 5),
                        breaks = c(0.001,0.05,0.1,0.2,0.4,0.6),
                        name = "Mean Abundance") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "right") + 
  facet_grid(Host~Year) +
  ggtitle("Indicator Genera") +
  xlab("Season") +
  ylab("")

ind_genera_plot_season

#and you can do so much more. these were just quick plots... could also do line plots with standard deviations

```

```{R indicators across wasting categories}

smic_gen.ind_ave <- tibble(smic_isa.genusm) %>% group_by(OTU,Host, Phylum, Class, Order, Family, Genus, Ind_status, Year, YearMonth, Season, SiteType, WastingScore, WastingCategory, DermisType2) %>% summarise(ave_abun = mean(Abundance)) 
#tidyr:: if unite doesn't work

smic_gen.ind_ave <-  smic_gen.ind_ave %>%
  mutate(DermisType2 = factor(DermisType2, levels = c("H_healthy", "H_diseased", "L_diseased")))


ind_genera_plot_wasting<-smic_gen.ind_ave  %>%
  tidyr::unite(GENIND, Genus, Ind_status, sep = "_", remove = FALSE) %>%
  filter(Host == "Pisaster ochraceus") %>%
  filter(Ind_status == "Pisaster") %>%
ggplot(aes(x = DermisType2, y = GENIND, size = ave_abun, color = Class)) +
  geom_point(alpha = 0.7) +
  scale_color_discrete_sequential(palette = "Viridis") +
  scale_size_continuous(range = c(0.5, 5),
                        breaks = c(0.001,0.05,0.1,0.2,0.4,0.6),
                        name = "Mean Abundance") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "right") + 
  facet_grid(SiteType~WastingCategory) +
  ggtitle("Indicator Genera") +
  xlab("Swab Type") +
  ylab("Genus-Indicator Group")

ind_genera_plot_wasting


#now let's look at a boxplot of the differences in abundance of these indicator groups across wasting categories, for example

smic_isa.genusm <- smic_isa.genusm %>%
  mutate(DermisType2 = factor(DermisType2, levels = c("H_healthy", "H_diseased", "L_diseased")))

#boxplot of genera by wasting cat and region
ind_genera_boxplot_wasting<-smic_isa.genusm %>%
  filter(Host == "Pisaster ochraceus") %>%
  filter(Ind_status == "Pisaster") %>%
ggplot(aes(x = DermisType2, y = log(Abundance), fill = DermisType2)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("#26547c", "#ffd166", "#ef476f"))+
  # scale_color_discrete_sequential(palette = "Mako") +
  scale_size_continuous(range = c(0.5, 5),
                        breaks = c(0.001,0.05,0.1,0.2,0.4,0.6),
                        name = "Mean Abundance") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.position = "right") + 
  facet_wrap(Genus~., scales = "free") +
  ggtitle("Indicator Genera") +
  xlab("Wasting Score") +
  ylab("Genus-Indicator Group")


ind_genera_boxplot_wasting

```


########### ASV LEVEL INDICATORS ---- not currently included   ########### 


```{r ASV level Indicator Species Analysis}

#### run indicator species analysis ####
host.indval = multipatt(smicR, hostgroups, control = how(nperm=999), duleg = TRUE)
#indval <- readRDS("multipatt_indval.RDS")

summary(host.indval)
attributes(host.indval)

isa.asv<-host.indval$sign

#write.csv(isa.asv, "SMIC-multipatt_ASV_indval_results.csv") #save the results

#### filter data frame for significant indicators ####
#In order to be included we want to retain only indicators with a p value of 0.001 or less and a stat of 0.7 or higher. I just picked these settings but we can adjust if need be.

isa_tbl<-tibble(rownames_to_column(isa.asv, var = "ASVID")) %>% rename(c("Seawater" = "s.Seawater", "Dermasterias" = "s.Dermasterias imbricata", "Pisaster" = "s.Pisaster ochraceus")) %>% mutate("ind_group"=index) %>%  mutate(ind_group = replace(ind_group, ind_group == 1, "Dermasterias")) %>% mutate(ind_group = replace(ind_group, ind_group == 2, "Pisaster")) %>% mutate(ind_group = replace(ind_group, ind_group == 3, "Seawater"))

#how many signficant indicator ASVs are there in your different sample groups (you can change the parameters for signficance)
sum(isa_tbl$index == 3 & isa_tbl$stat >= 0.7 & isa_tbl$p.value <= 0.001) 
#[1] 134 = seawater indicators

sum(isa_tbl$index == 2 & isa_tbl$stat >= 0.7 & isa_tbl$p.value <= 0.001)
#[1] 13= pisaster indicators

sum(isa_tbl$index == 1 & isa_tbl$stat >= 0.7 & isa_tbl$p.value <= 0.001)
#[1] 8 = derm indicators

#make a new column denoting indicator status

isa_tbl <- isa_tbl %>% mutate("isInd" = case_when(
  stat >= 0.7 & p.value <= 0.001 ~ "Yes",
  stat < 0.7 | p.value > 0.001 ~ "No"
  ))

isa_tbl <-isa_tbl %>% mutate("Ind_status" = ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Seawater", 'Seawater', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Pisaster", 'Pisaster', ifelse(stat >= 0.7 & p.value <= 0.001 & ind_group == "Dermasterias", 'Dermasterias', 'No'))))

isa_tbl


#### merge indicator status with taxonomy table ####

#now we need to add a column about indicator status to the taxonomy table

Ind_tbl <- isa_tbl %>% select(ASVID, Ind_status)
Ind_tbl

### map back to the full data frame, including diseased stars ###

smicpf_taxa<-data.frame(tax_table(smicpf_sorted)) %>% rownames_to_column() %>% 
  dplyr::rename(ASVID = "rowname") 

smic_species_tax_ind<-left_join(smicpf_taxa, Ind_tbl)

smic_species_tax_ind<-as.matrix(column_to_rownames(as.data.frame(smic_species_tax_ind), var = "ASVID"))

#### update taxonomy table in phyloseq object ####
smic_isa.asv <- smicpf_sorted #rename original phyloseq object
tax_table(smic_isa.asv) <- smic_species_tax_ind #update taxa table to include ISA

#write.csv(smic_species_tax_ind, "smic_species_tax_ind.csv")

```

