---
title: "compare taxa means"
author: "Colleen Kellogg"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}

library(phyloseq)
library(microViz)
library(speedyseq)
library(ggpubr)
library(rstatix)
library(dplyr)
#library(microbiomeutilities)
library(microbiome)
library(tidyverse)
library(kableExtra)
library(tidyr)

```

## Load data and determine top genera
```{R load data and determine top genera}
#### prep phyloseq ####
smicpf <- readRDS("C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/decontam_filter_nonegs_prev.rds")

pis_gen <- smicpf %>%
  transform("compositional") %>%
  aggregate_taxa("Genus") %>%
  subset_samples(Host == "Pisaster ochraceus")

pis_genm <- speedyseq::psmelt(pis_gen)
 pis_genm

#### get top genera ####
top_12g <- tax_top(pis_gen, n = 12, rank = "Genus")
top_15g <- tax_top(pis_gen, n = 15, rank = "Genus")
top_20g <- tax_top(pis_gen, n = 20, rank = "Genus")
top_50g <- tax_top(pis_gen, n = 50, rank = "Genus") 

top_50g
```

## Run wilcoxon test to compare means

I went with 0.01 because BH is a more permissive multiple test correction. with holm @ 0.05, only 11/50 genera were significantly different. but with BH at 0.05, 36/50 were sig different and in some cases, they had minimal log2FC, with 0.01 it seems like it is a nice compromise, with 22/50 top genera showing significant differences that looks visibly different too. this can obviously be changed!

```{R run wilcoxon test}
#### stats — wilcox with BH correction ####
#if you want to save the stats in a CSV file, do this.
stats_results_category <- pis_genm %>%
  filter(Genus %in% top_50g) %>%
  group_by(Genus) %>%
  wilcox_test(Abundance ~ WastingCategory) %>% #ref.group = "Healthy" -- by setting a ref.group that will only calculate and plot stats for in reference to that group (rather than also including severe vs mid); remove ref.group if you want to know about those too.
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

#write_csv(stats_results_category,"SMIC-top50genera_wilcoxon_stats_results.csv")

#to look at the results in table format:
stats_results_category %>%
  kable(
    format   = "html",
    digits   = 3,
    caption  = "Results of comparison (wilcoxon test) of healthy, mildly and severely wasting pisaster sea stars",
    # col.names = c("Genus", "value compared", "group1", "group2", "n1", "n2","statistic","p","p.adj","p.adj.signif")   # rename cols
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width        = FALSE,
    position          = "center",
    font_size         = 12
  ) %>%
  row_spec(0, bold = TRUE) %>%           # bold header row
  column_spec(1, italic = TRUE)


#### significant genera with at least one significant comparison vs Healthy ####

sig_genera <- stats_results_category %>%
  filter(p.adj < 0.01) %>% 
  pull(Genus) %>%
  unique()

cat("Significant genera vs Healthy:", length(sig_genera), "\n")
print(sig_genera)

#### subset significant genera ####
siggenera_to_plot <- pis_genm %>%
  filter(Genus %in% sig_genera) %>%
  mutate(WastingCategory = factor(WastingCategory, 
                                  levels = c("Healthy", "Mild", "Severe")))


```

## A few ways to plot Wilcoxon test to compare group means

```{R ways to visualize compare means results}
#### plot ####
library(scales)
library(ggpubr)
        
#plot using ggpubr and geom_pwc, which employs the same wilcoxon test as wilcox_test in rstatix, so the results should be the same

wastingcategory_boxplot <- ggboxplot(
    siggenera_to_plot,
    y = "Abundance",
    x = "WastingCategory",
    fill = "WastingCategory",
    lwd = 0.3,
    alpha = 0.5
  ) + 
  geom_pwc(
    aes(group = WastingCategory),
    tip.length = 0.01,
    method = "wilcox_test",
    ref.group = "Healthy",
    p.adjust.method = "BH",
    p.adjust.by = "group",
    label = "p.adj.signif",         # or "p.adj" for numeric values
    step.increase = 0.1,
    size = 0.2,
    label.size = 3,
    vjust = 0.75,
    hide.ns = TRUE
  ) +
  scale_fill_manual(
    values = c("lightseagreen", "mediumorchid1", "mediumorchid4"),
    name = "Wasting Category"
  ) +
  scale_y_sqrt() +
  #scale_y_log10(breaks = c(0.0001, 0.001, 0.01, 0.1, 1)) + 
  facet_wrap(Genus ~., nrow = 4,) +
  theme_classic() +
  theme(
    axis.text.x     = element_blank(),
    axis.title.x    = element_blank(),
    legend.position = "bottom",
    strip.placement = "outside",
    strip.text = element_text(size = 6),
    strip.text.x.bottom = element_text(angle = 90, hjust = 1, size = 8),
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ylab("Relative Abundance")

wastingcategory_boxplot

ggsave("SMIC-wastingcategory_boxplot.pdf",wastingcategory_boxplot, height = 6, width = 8)



#### calculate log fold change ####


SMIC_siggenera_WC_log2FC<-pis_genm %>% group_by(Genus, WastingCategory) %>% summarise(MeanAbun = mean(Abundance)) %>% tidyr::pivot_wider(id_cols = Genus, values_from = MeanAbun, names_from = WastingCategory) %>% mutate(log2FC_HM = log2(Mild/Healthy), log2FC_HS = log2(Severe/Healthy), log2FC_MS = log2(Severe/Mild)) %>% filter(Genus %in% sig_genera)

#### plot log2(fold change) bar plots ####

# mild:healthy
SMIC_mildhealthy_log2FC<-SMIC_siggenera_WC_log2FC %>%
ggplot(aes(x = log2FC_HM, y = reorder(Genus, log2FC_HM))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_bw() + 
  ylab(NULL) +
  xlab ("Log2(Fold Change of mildly-wasting:healthy sea stars)")

SMIC_mildhealthy_log2FC

#severe:healthy
SMIC_severehealthy_log2FC<-SMIC_siggenera_WC_log2FC %>%
ggplot(aes(x = log2FC_HS, y = reorder(Genus, log2FC_HS))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_bw() + 
  ylab(NULL) +
  xlab ("Log2(Fold Change of severe-wasting:healthy sea stars)")

SMIC_severehealthy_log2FC

#### pivot log2FC data frame longer for plotting ###

SMIC_siggenera_WC_log2FC.long<-SMIC_siggenera_WC_log2FC %>% pivot_longer(
  cols = c(log2FC_HS,log2FC_HM, log2FC_MS), names_to = "abun_comp", values_to = "Log2FC"
) %>% mutate(abun_comp = recode(abun_comp,
    "log2FC_HM" = "Mild:Healthy",
    "log2FC_HS" = "Severe:Healthy",
    "log2FC_MS" = "Severe:Mild"
  ))

#### log2FC paired barplot ####

SMIC_severehealthy_log2FC.bar<-SMIC_siggenera_WC_log2FC.long %>%
ggplot(aes(x = Log2FC, y = reorder(Genus,Log2FC), fill = abun_comp)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#ffd166", "#ef476f", "coral1"), name = "Comparison")+
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_classic() + 
  ylab(NULL) +
  xlab ("Log2 Fold Change")

SMIC_severehealthy_log2FC.bar #need to figure out how to indicate significance


#### log2 fold change faceted lollipop plot ####

SMIC_severehealthy_log2FC.dots <- SMIC_siggenera_WC_log2FC.long %>%
  ggplot(aes(x = Log2FC, y = reorder(Genus, Log2FC), color = abun_comp)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_segment(
    aes(x = 0, xend = Log2FC,
        y = reorder(Genus, Log2FC), 
        yend = reorder(Genus, Log2FC)),
    linewidth = 0.4, alpha = 0.6        # lollipop stem
  ) +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  scale_color_manual(values = c("#ffd166", "#ef476f", "coral1"), name = "Comparison") +
  theme_classic() +
  ylab(NULL) +
  xlab("Log2 Fold Change") +
  facet_grid(~abun_comp) +
  theme(strip.background = element_blank()) +
  theme(legend.position = "none")

SMIC_severehealthy_log2FC.dots

# can we add an asterix if the difference is significant?

#### join log2 fold change and wilcoxon results ####

stats_results_summary <- stats_results_category %>%
  unite(abun_comp, group2, group1, sep = ":", remove = FALSE)  %>% select(Genus, abun_comp, p.adj, p.adj.signif)


SMIC_siggenera_WC_log2FC_sig<-left_join(SMIC_siggenera_WC_log2FC.long,stats_results_summary)

SMIC_siggenera_WC_log2FC_sig

#### lollipop plots with significance ####

SMIC_siggenera_WC_log2FC_sig %>%
  mutate(sig_label = ifelse(p.adj < 0.05, "*", "")) %>%   # or your significance column
  ggplot(aes(x = Log2FC, y = reorder(Genus, Log2FC), color = abun_comp)) +
  geom_segment(
    aes(x = 0, xend = Log2FC,
        y = reorder(Genus, Log2FC),
        yend = reorder(Genus, Log2FC)),
    linewidth = 0.4, alpha = 0.6
  ) +
  geom_point(size = 4, alpha = 0.85) +
  geom_text(aes(label = sig_label),
            color = "white",              # white asterisk inside dot
            size = 4,
            fontface = "bold",
            vjust = 0.75 ) +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  # scale_color_manual(values = c("#ffd166", "#ef476f"), name = "Comparison")
  scale_color_manual(values = c("#ffd166", "#ef476f", "coral1"), )+
  theme_minimal() +
  theme(legend.position = "none") +
  ylab(NULL) +
  xlab("Log2 Fold Change") +
  facet_grid(~abun_comp)

#### bar plot with significance ####
logchange_category<-SMIC_siggenera_WC_log2FC_sig %>%
  mutate(sig_label = recode(p.adj.signif, 'ns' = "")) %>%   # or your s
  ggplot(aes(x = Log2FC, y = reorder(Genus, Log2FC), fill = abun_comp)) +
  geom_col(position = "dodge") +
  # stars at end of each bar
  geom_text(
    aes(label = sig_label,
        x = ifelse(Log2FC >= 0,       # place star at end of bar
                   Log2FC + 0.05,     # positive bars: right of bar
                   Log2FC - 0.05),    # negative bars: left of bar
        hjust = ifelse(Log2FC >= 0, 0, 1),
        vjust = 0.75),
    position = position_dodge(width = 0.9),
    color = "grey20",
    size = 3.5
  ) +
  scale_fill_manual(values = c("#ffd166", "#ef476f", "coral1"), name = "Comparison") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylab(NULL) +
  xlab("Log2 Fold Change") + 
  facet_grid(~abun_comp) +
  theme(strip.background = element_blank())  + 
  ggtitle("Wasting Category Comparison")

logchange_category

```


#### RUNNING THE SAME ANALYIS BUT WITH TISSUE TYPE INSTEAD OF WASTING CATEGORY   ####

```{r}
#### stats — wilcox with BH correction ####

stats_results_tissue <- pis_genm %>%
  filter(Genus %in% top_50g) %>%
  group_by(Genus) %>%
  wilcox_test(Abundance ~ DermisType2) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

stats_results_tissue

sig_genera_tissue <- stats_results_tissue %>%
  filter(p.adj < 0.01) %>% 
  pull(Genus) %>%
  unique()

cat("Significant genera vs H_healthy:", length(sig_genera_tissue), "\n")

print(sig_genera_tissue)


#### subset significant genera ####
siggenera_to_plot_tissue <- pis_genm %>%
  filter(Genus %in% sig_genera_tissue) %>%
  mutate(DermisType2 = factor(DermisType2, 
                                  levels = c("H_healthy", "H_diseased", "L_diseased")))

```

#now trying plotting across tissue types

```{r}

tissuetype_boxplot <- ggboxplot(
    siggenera_to_plot_tissue,
    y = "Abundance",
    x = "DermisType2",
    fill = "DermisType2",
    lwd = 0.3,
    alpha = 0.7
  ) +
  geom_pwc(
    aes(group = DermisType2),
    tip.length = 0.01,
    method = "wilcox_test",
    ref.group = "H_healthy",
    p.adjust.method = "BH",
    p.adjust.by = "group",
    label = "p.adj.signif",         # or "p.adj" for numeric values
    step.increase = 0.1,
    size = 0.2,
    label.size = 3,
    vjust = 0.75,
    hide.ns = TRUE
  ) +
  scale_fill_manual(
    values = c("olivedrab", "deepskyblue", "deepskyblue4"),
    name = "Tissue Type"
  ) +
  scale_y_sqrt() +
  # scale_y_log10(breaks = c(0.0001, 0.001, 0.01, 0.1, 1),    
  facet_wrap(Genus ~., nrow = 4,) +
  theme_bw() +
  theme(
    axis.text.x     = element_blank(),
    axis.title.x    = element_blank(),
    legend.position = "bottom",
    strip.placement = "outside",
    strip.text = element_text(size = 6),
    strip.text.x.bottom = element_text(angle = 90, hjust = 1, size = 8),
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ylab("Relative Abundance")

tissuetype_boxplot


```


```{r}
#### calculate log fold change ####

SMIC_siggenera_TT_log2FC<-pis_genm %>%
  group_by(Genus, DermisType2) %>%
  summarise(MeanAbun = mean(Abundance)) %>%
  pivot_wider(id_cols = Genus, values_from = MeanAbun, names_from = DermisType2) %>%
  mutate(log2FC_HH = log2(H_diseased/H_healthy), log2FC_HL1 = log2(L_diseased/H_diseased), log2FC_HL2 = log2(L_diseased/H_healthy), ) %>% filter(Genus %in% sig_genera) 

SMIC_siggenera_TT_log2FC


#### pivot log2FC data frame longer for plotting ###

SMIC_siggenera_TT_log2FC.long<-SMIC_siggenera_TT_log2FC %>% pivot_longer(
  cols = c(log2FC_HH,log2FC_HL1, log2FC_HL2), names_to = "abun_comp", values_to = "Log2FC"
) %>% mutate(abun_comp = recode(abun_comp,
    "log2FC_HH" = "H_diseased:H_healthy",
    "log2FC_HL1" = "L_diseased:H_diseased", 
    "log2FC_HL2" = "L_diseased:H_healthy"
  ))

SMIC_siggenera_TT_log2FC.long

#### log2FC paired barplot ####

SMIC_tissuetype_log2FC.bar<-SMIC_siggenera_TT_log2FC.long %>%
ggplot(aes(x = Log2FC, y = reorder(Genus,Log2FC), fill = abun_comp)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("firebrick", "mediumpurple1", "cyan4"), name = "Comparison")+
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_classic() + 
  ylab(NULL) +
  xlab ("Log2 Fold Change")


```

```{r}

#### join log2 fold change and wilcoxon results ####

stats_results_summary_tissue <- stats_results_tissue %>%
  unite(abun_comp, group2, group1, sep = ":", remove = FALSE)  %>% select(Genus, abun_comp, p.adj, p.adj.signif)

stats_results_summary_tissue

SMIC_siggenera_TT_log2FC_sig<-left_join(SMIC_siggenera_TT_log2FC.long,stats_results_summary_tissue)


SMIC_siggenera_TT_log2FC_sig

#blues
SMIC_siggenera_TT_log2FC_sig %>%
  mutate(sig_label = ifelse(p.adj < 0.05, "*", "")) %>%   # or your significance column
  ggplot(aes(x = Log2FC, y = reorder(Genus, Log2FC), color = abun_comp)) +
  geom_segment(
    aes(x = 0, xend = Log2FC,
        y = reorder(Genus, Log2FC),
        yend = reorder(Genus, Log2FC)),
    linewidth = 0.4, alpha = 0.6
  ) +
  geom_point(size = 4, alpha = 0.85) +
  geom_text(aes(label = sig_label),
            color = "white",              # white asterisk inside dot
            size = 4,
            fontface = "bold",
            vjust = 0.75 ) +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  # scale_color_manual(values = c("#ffd166", "#ef476f"), name = "Comparison")
  scale_color_manual(values = c("#84b0a9","#1f3e74", "mediumpurple1"), )+
  theme_minimal() +
  theme(legend.position = "none") +
  ylab(NULL) +
  xlab("Log2 Fold Change") +
  facet_grid(~abun_comp)


#### bar plot with significance ####
logchange_tissue<-SMIC_siggenera_TT_log2FC_sig %>%
  mutate(sig_label = recode(p.adj.signif, 'ns' = "")) %>%   # or your s
  ggplot(aes(x = Log2FC, y = reorder(Genus, Log2FC), fill = abun_comp)) +
  geom_col(position = "dodge") +
  # stars at end of each bar
  geom_text(
    aes(label = sig_label,
        x = ifelse(Log2FC >= 0,       # place star at end of bar
                   Log2FC + 0.05,     # positive bars: right of bar
                   Log2FC - 0.05),    # negative bars: left of bar
        hjust = ifelse(Log2FC >= 0, 0, 1),
        vjust = 0.75),
    position = position_dodge(width = 0.9),
    color = "grey20",
    size = 3.5
  ) +
  scale_fill_manual(values = c("firebrick", "mediumpurple1", "cyan4"), name = "Comparison") +
  geom_vline(xintercept = 0, linewidth = 0.5, color = "black") +
  theme_minimal() +
  theme(legend.position = "none") +
  ylab(NULL) +
  xlab("Log2 Fold Change") + 
  facet_grid(~abun_comp) + 
  ggtitle("Tissue Type Comparison")

logchange_tissue

```


#if we want to combine both log change plots

```{r}

logchange_panel<-ggarrange(logchange_category, logchange_tissue, ncol=1, align = "v", legend = "right", labels = c("a", "b"))

logchange_panel

ggsave("logchange_panel.pdf", dpi=300, width=10, height=8, units="in")

```


