---
title: "smic_beta_diversity"
output: html_document
date: "2026-03-12"
---

########## BETA DIVERSITY ##############

First read in the phyloseq object that we need

```{r }

#read in the required QC'ed phyloseq object
decontam_filter_nonegs_prev <- readRDS("C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/decontam_filter_nonegs_prev.rds")

```

```{r load required packages}

library(phyloseq)
library(ggplot2)
library(vegan)
library(ggpubr)
library(RColorBrewer)
library(microViz)

```


Beta diversity among host and regions

```{r Ordinations}

decontam_filter_nonegs_prev

#for the host comparison we just want healthy pisaster along with derm and seawater
healthy_all_b <- ps_filter(
  decontam_filter_nonegs_prev,
  (Host == "Pisaster ochraceus" & DermisType2 == "H_healthy") |
  (Host == "Dermasterias imbricata" & DermisType2 == "H_healthy") |
  (Host == "Seawater"))

#first we need to normalize the data, regardless of ordination method

healthy_all_b_norm <- healthy_all_b %>%
  transform_sample_counts(function(x) {x/sum(x)})

#note, tried NMDS here, including increasing trymax = 1000 and k=3, also tried with Hellinger transformation and was not converging/stress values were on the cusp of acceptable so decided to go with PCoA

healthy_host_pcoa <- phyloseq::ordinate(healthy_all_b_norm, method = "PCoA", distance = "bray")


pcoa_sampletype <-plot_ordination(healthy_all_b_norm, healthy_host_pcoa, type = "samples",
  color = "Host", shape="SiteType") +
  theme_classic() +
   scale_color_manual(values = c("#AD002AFF", "#925E9FFF","#0099B4FF"), 
   labels = c(expression(italic("D. imbricata")),
   expression(italic("P. ochraceus")),"Seawater")) + 
  scale_shape_manual(values = c(8, 15)) +
  theme(
    legend.position = "right",
    strip.background = element_blank()) + 
  labs(color = "Sample Type", shape = "Region")


pcoa_sampletype


```

Combining alpha and diversity plots by sample type to create Figure 2

```{r}

alpha_plot <- readRDS("C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/alpha_host_boxplot.rds")

fig2<-ggarrange(alpha_plot, pcoa_sampletype, legend = "right", labels = c("A", "B"), ncol=1, heights = c(1, 1.5), common.legend = FALSE)

fig2

#saving locally to computer
ggsave(plot = fig2, filename = file.path("C:/Users/Carolyn/Desktop/Paper Figures","host_alpha_beta.jpeg"), dpi=300, width=9, height=8, units="in")


#IF wanting to save to github repo eventually

#ggsave(plot = fig2, filename = file.path("C:/Users/Carolyn/Documents/seastar-microbiome-paper/figures","host_alpha_beta.jpeg"), dpi=300, width=9, height=8, units="in")


```



```{r permanova to look at effect of host/sample type on alpha div}

#using adonis to run a PERMANOVA by host (aka sample type)

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_host <- phyloseq::distance(healthy_all_b_norm, method = "bray")

# make a data frame from the sample_data
host_df <- data.frame(sample_data(healthy_all_b_norm))

# Permanova by Host
adonis2(dist_mat_host~ Host*SiteType, data = host_df, permutations = 999)


#beta dispersion test

bd_host <- betadisper(dist_mat_host, interaction(host_df$Host, host_df$SiteType))
anova(bd_host)

# P < 0.001, significant value indicates that the dispersions are NOT homogeneous among groups...we can visualize this
plot(bd_host)

boxplot(bd_host, xlab = "Sample Type", ylab = "Distance to centroid")

```

 
```{r Testing for effect of region within seawater}

seawater_norm <- ps_filter(healthy_all_b_norm,
  (Host == "Seawater"))

seawater.pcoa <- ordinate(seawater_norm, method = "PCoA", distance = "bray")

# Plot of seawater communities between regions
seawater_region_ord<-plot_ordination(
  physeq = seawater_norm,
  ordination = seawater.pcoa,
  color = "SiteType", shape = "Season") +
  scale_color_manual(values = c("orange","mediumturquoise")) + 
  theme_classic() +
   ggtitle("Seawater")

seawater_region_ord

#using adonis to run a PERMANOVA by region within seawater

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_seawater <- phyloseq::distance(seawater_norm, method = "bray")

# make a data frame from the sample_data
seawater_df <- data.frame(sample_data(seawater_norm))

# Permanova by region within seawater
adonis2(dist_mat_seawater~ SiteType, data = seawater_df, permutations = 999)

#beta dispersion test

bd_sw_region <- betadisper(dist_mat_seawater, seawater_df$SiteType)
anova(bd_sw_region)   #p =0.1845 so meets homogeneity of dispersion assumptions

plot(bd_sw_region)

```

 
```{r Testing for effect of region within Derm}

derm_norm <- ps_filter(healthy_all_b_norm,
  (Host == "Dermasterias imbricata"))

derm.pcoa <- ordinate(derm_norm, method = "PCoA", distance = "bray")

# Plot of seawater communities between regions
derm_region_ord<-plot_ordination(
  physeq = derm_norm,
  ordination = derm.pcoa,
  color = "SiteType") +
  scale_color_manual(values = c("orange","mediumturquoise")) + 
  theme_classic() +
   ggtitle("D. imbricata") + theme(plot.title = element_text(size = 10, face = "italic"))

derm_region_ord

#using adonis to run a PERMANOVA by region within Derm

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_derm <- phyloseq::distance(derm_norm, method = "bray")

# make a data frame from the sample_data
derm_df <- data.frame(sample_data(derm_norm))

# Permanova by region with Derm samples
adonis2(dist_mat_derm~ SiteType, data = derm_df)

#beta dispersion test
bd_derm_region <- betadisper(dist_mat_derm, derm_df$SiteType)
anova(bd_derm_region) # P = 0.1315 so meets homogeneity of dispersion assumptions


plot(bd_derm_region)

```


```{r Testing for effect of region within Pisaster}

pis_norm <- ps_filter(healthy_all_b_norm,
  (Host == "Pisaster ochraceus"))

pis.pcoa <- ordinate(pis_norm, method = "PCoA", distance = "bray")

# Plot of seawater communities between regions
pis_region_ord<-plot_ordination(
  physeq = pis_norm,
  ordination = pis.pcoa,
  color = "SiteType") +
  scale_color_manual(values = c("orange","mediumturquoise")) + 
  theme_classic() +
   ggtitle("P. ochraceus") +theme(plot.title = element_text(size = 10, face = "italic"))

pis_region_ord

#using adonis to run a PERMANOVA by region within Pisaster

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_pis <- phyloseq::distance(pis_norm, method = "bray")

# make a data frame from the sample_data
pis_df <- data.frame(sample_data(pis_norm))

# Permanova by region with Derm samples
adonis2(dist_mat_pis~ SiteType, data = pis_df)

#beta dispersion test
bd_pis_region <- betadisper(dist_mat_pis, pis_df$SiteType)
anova(bd_pis_region) # P < 0.001 so does not meet homogeneity of dispersion assumption


plot(bd_pis_region)


```

############ SEASONAL TRENDS IN BETA DIVERSITY #############
 
 
```{r Seasonal trends in beta diversity in seawater}

# Plot of seawater communities across seasons 
seawater_season_ord<-plot_ordination(
  physeq = seawater_norm,
  ordination = seawater.pcoa,
  color = "Season", shape = "Year") +
  #scale_color_manual(values = c("coral2", "lightseagreen", "darkgoldenrod2", "darkslateblue")) 
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) +
 facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
   ggtitle("Seawater") + theme(strip.background = element_blank() # This removes the box and background
  )

seawater_season_ord

#using adonis to run a PERMANOVA by season within seawater

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_seawater <- phyloseq::distance(seawater_norm, method = "bray")

# make a data frame from the sample_data
seawater_df <- data.frame(sample_data(seawater_norm))

# Permanova by season within seawater
adonis2(dist_mat_seawater~ Season, data = seawater_df)

#beta dispersion test
bd_sw_season <- betadisper(dist_mat_seawater, seawater_df$Season)
anova(bd_sw_season) # P = 0.0046 so does not meet homogeneity of dispersion assumption

plot(bd_sw_season)


```


```{r Derm beta diversity season}

# Plot of healthy Derm across seasons
derm_season_ord<-plot_ordination(
  physeq = derm_norm,
  ordination = derm.pcoa,
  color = "Season", shape = "Year") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) +
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
   ggtitle("D. imbricata") + theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank() # This removes the box and background
  )

derm_season_ord

#using adonis to run a PERMANOVA by season within Derm

# Permanova by Host
adonis2(dist_mat_derm~Season, data = derm_df)

#beta dispersion test
bd_derm_season<-betadisper(dist_mat_derm, derm_df$Season)
anova(bd_derm_season) #p = 0.00011 so does not meet homogeneity of dispersion assumption

plot(bd_derm_season)


```


```{r}

# Plot of healthy Pisaster across seasons
pis_season_ord<-plot_ordination(
  physeq = pis_norm,
  ordination = pis.pcoa,
  color = "Season", shape = "Year") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) +
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
   ggtitle("P. ochraceus") + theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank() # This removes the box and background
  )

pis_season_ord

#using adonis to run a PERMANOVA by season within Pisaster

# Permanova by season within pisaster
adonis2(dist_mat_pis ~ Season, data = pis_df)

#beta dispersion test
beta_pis_season <- betadisper(dist_mat_pis, pis_df$Season)
permutest(beta_pis_season) 

anova(beta_pis_season) #p = 0.02152 so does not meet homogeneity of dispersion assumption (but is close, would just need to be > 0.05)

plot(beta_pis_season)

```


Combining beta diversity plots
```{r}

season_panel_beta<-ggarrange(seawater_season_ord, derm_season_ord, pis_season_ord, common.legend=TRUE, ncol=3, align = "v", legend = "right", labels = c("D", "E", "F"))

readRDS("C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/season_panel_alpha.rds")

season_panel_alpha
season_panel_beta

season_alpha_beta<-ggarrange(season_panel_alpha, season_panel_beta, common.legend=TRUE, ncol=1, align = "v", legend = "right")

season_alpha_beta

#saving locally to computer
ggsave(plot = season_alpha_beta, filename = file.path("C:/Users/Carolyn/Desktop/Paper Figures","season_alpha_beta.jpeg"), dpi=300, width=9, height=8, units="in")


#ggsave(plot = season_alpha_beta,file.path("C:/Users/Carolyn/Documents/seastar-microbiome-paper/figures","season_alpha_beta.jpeg"), dpi=300, width=9, height=8, units="in")

```


############ SEASONAL TRENDS IN BETA DIVERSITY WITHOUT or ONLY CORE TAXA #############

```{R Derm beta diversity season NO CORE }
# DERM

derm_core<-c("Saprospiraceae Family","EC94", "Rhodobacteraceae Family","Flavobacteriaceae Family","Rubritalea")

derm_norm_nocore <- ps_filter(healthy_all_b_norm,
  (Host == "Dermasterias imbricata")) %>%
  subset_taxa(!Genus %in% derm_core)

derm.nocore.pcoa <- ordinate(derm_norm_nocore, method = "PCoA", distance = "bray")

# Plot of healthy Derm across seasons
derm_season_ord.nocore<-plot_ordination(
  physeq = derm_norm_nocore,
  ordination = derm.nocore.pcoa,
  color = "Season",
  shape = "SiteType") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) + 
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
  ggtitle("D. imbricata") + 
  theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank()) # This removes the box and background

derm_season_ord.nocore

#using adonis to run a PERMANOVA by season within Derm, without core

# make a data frame from the sample_data
derm_df.nocore <- data.frame(sample_data(derm_norm_nocore))

dist_mat_derm.nocore <- phyloseq::distance(derm_norm_nocore, method = "bray")

# Permanova by Host
adonis2(dist_mat_derm.nocore~Season, data = derm_df.nocore) #p = 0.001

#beta dispersion test
bd_derm_season.nocore<-betadisper(dist_mat_derm.nocore, derm_df.nocore$Season)
anova(bd_derm_season.nocore) #p = 2.909e-05 

plot(bd_derm_season.nocore)


```


```{R Derm beta diversity season ONLY CORE }
# DERM

derm_core<-c("Saprospiraceae Family","EC94", "Rhodobacteraceae Family","Flavobacteriaceae Family","Rubritalea")

derm_norm_core <- ps_filter(healthy_all_b_norm,
  (Host == "Dermasterias imbricata")) %>%
  subset_taxa(Genus %in% derm_core)

derm.core.pcoa <- ordinate(derm_norm_core, method = "PCoA", distance = "bray")

# Plot of healthy Derm across seasons
derm_season_ord.core<-plot_ordination(
  physeq = derm_norm_core,
  ordination = derm.core.pcoa,
  color = "Season",
  shape = "SiteType") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) + 
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
  ggtitle("D. imbricata") + 
  theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank()) # This removes the box and background

derm_season_ord.core

#using adonis to run a PERMANOVA by season within Derm, without core

# make a data frame from the sample_data
derm_df.core <- data.frame(sample_data(derm_norm_core))

dist_mat_derm.core <- phyloseq::distance(derm_norm_core, method = "bray")

# Permanova by Host
adonis2(dist_mat_derm.core~Season, data = derm_df.core) #p = 0.001

#beta dispersion test
bd_derm_season.core<-betadisper(dist_mat_derm.core, derm_df.core$Season)
anova(bd_derm_season.core) #p = 0.0002842

plot(bd_derm_season.core)

```


```{R Pisaster beta diversity season NO CORE}
#PISASTER

pioc_core <-c("Roseibacillus","Polaribacter","Yoonia-Loktanella",       "Rubritalea","Saprospiraceae Family","Reichenbachiella",        "Endozoicomonas","Rhodobacteraceae Family","Spirochaetia Class", "Spirochaetaceae Family","Salinispira","Flavobacteriaceae Family")

pioc_norm_nocore <- ps_filter(healthy_all_b_norm,
  (Host == "Pisaster ochraceus")) %>%
  subset_taxa(!Genus %in% pioc_core)

# pis.nocore.pcoa <- ordinate(pis_norm, method = "NMDS", distance = "bray")
pioc.nocore.pcoa <- ordinate(pioc_norm_nocore, method = "PCoA", distance = "bray")

# Plot of healthy Pisaster across seasons
pioc_season_ord.nocore<-plot_ordination(
  physeq = pioc_norm_nocore,
  ordination = pioc.nocore.pcoa,
  color = "Season",
  shape = "SiteType") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) + 
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
  ggtitle("P. ochraceus") + 
  theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank()) # This removes the box and background

pioc_season_ord.nocore

### permanova and anosim ###

#using adonis to run a PERMANOVA by season within Derm, without core

# make a data frame from the sample_data
pioc_df.nocore <- data.frame(sample_data(pioc_norm_nocore))

dist_mat_pioc.nocore <- phyloseq::distance(pioc_norm_nocore, method = "bray")

# Permanova by season
adonis2(dist_mat_pioc.nocore~Season, data = pioc_df.nocore) #p = 0.001

#beta dispersion test
bd_pioc_season.nocore<-betadisper(dist_mat_pioc.nocore, pioc_df.nocore$Season)
anova(bd_pioc_season.nocore) #p = 7.485e-06

plot(bd_pioc_season.nocore)


```

```{R Pisaster beta diversity season ONLY CORE}
#PISASTER

pioc_core <-c("Roseibacillus","Polaribacter","Yoonia-Loktanella",       "Rubritalea","Saprospiraceae Family","Reichenbachiella",        "Endozoicomonas","Rhodobacteraceae Family","Spirochaetia Class", "Spirochaetaceae Family","Salinispira","Flavobacteriaceae Family")

pioc_norm_core <- ps_filter(healthy_all_b_norm,
  (Host == "Pisaster ochraceus")) %>%
  subset_taxa(Genus %in% pioc_core)

# pis.nmds <- ordinate(pis_norm, method = "NMDS", distance = "bray")
pioc.core.pcoa <- ordinate(pioc_norm_core, method = "PCoA", distance = "bray")

# Plot of healthy Pisaster across seasons
pioc_season_ord.core<-plot_ordination(
  physeq = pioc_norm_core,
  ordination = pioc.core.pcoa,
  color = "Season",
  shape = "SiteType") +
  scale_color_manual(values=c("#FE5E41", "#00A896","#9DD9D2", "#05668D")) + 
  facet_wrap(~SiteType, axes = "all_x", ncol=1, scales="free") +
  theme_classic() +
  ggtitle("P. ochraceus") + 
  theme(plot.title = element_text(size = 10, face = "italic")) + theme(strip.background = element_blank()) # This removes the box and background

pioc_season_ord.core

### permanova and anosim ###

#using adonis to run a PERMANOVA by season within Derm, without core

# make a data frame from the sample_data
pioc_df.core <- data.frame(sample_data(pioc_norm_core))

dist_mat_pioc.core <- phyloseq::distance(pioc_norm_core, method = "bray")

# Permanova by Host
adonis2(dist_mat_pioc.core~Season, data = pioc_df.core) #p = 0.001

#beta dispersion test
bd_pioc_season.core<-betadisper(dist_mat_pioc.core, pioc_df.core$Season)
anova(bd_pioc_season.core) #p = 7.485e-06

plot(bd_pioc_season.core)

```



############ WASTING DISEASE TRENDS IN BETA DIVERSITY #############


```{r Pisaster wasting scores beta diversity}

#want all Pisaster in this case, not just healthy

pisaster_all <- ps_filter(decontam_filter_nonegs_prev,
  (Host == "Pisaster ochraceus")) 

pisaster_quadra <- ps_filter(decontam_filter_nonegs_prev,
  (Host == "Pisaster ochraceus" & SiteType == "Quadra")) 

pisaster_calvert <- ps_filter(decontam_filter_nonegs_prev,
  (Host == "Pisaster ochraceus" & SiteType == "Calvert")) 

  
pisaster_quadra_norm <- pisaster_quadra %>%
  transform_sample_counts(function(x) {x/sum(x)})

pisaster_calvert_norm <- pisaster_calvert %>%
  transform_sample_counts(function(x) {x/sum(x)})

table(sample_data(pisaster_calvert_norm)$WastingCategory)
table(sample_data(pisaster_quadra_norm)$DermisType2)
table(sample_data(pisaster_quadra_norm)$Disease_Status_Star)


#ordination with bray and PCoA, within each region separately
diseased_pisaster_quadra_pcoa <- ordinate(pisaster_quadra_norm, 
    method = "PCoA", distance = "bray")

#ordination with bray and PCoA
diseased_pisaster_calvert_pcoa <- ordinate(pisaster_calvert_norm, 
    method = "PCoA", distance = "bray")


# Plot of Pisaster based on healthy vs diseased stars
pisaster_diseasestat_quadra <-plot_ordination(
physeq = pisaster_quadra,
  ordination = diseased_pisaster_quadra_pcoa,
  color = "Disease_Status_Star") +
 scale_color_manual(values = c("turquoise2", "maroon")) + 
  facet_grid(Year~Site) +
  theme_classic()

pisaster_diseasestat_quadra

pisaster_diseasestat_calvert <-plot_ordination(
physeq = pisaster_calvert,
  ordination = diseased_pisaster_calvert_pcoa,
  color = "Disease_Status_Star") +
 scale_color_manual(values = c("turquoise2", "maroon")) + 
  facet_grid(Year~Site) +
  theme_classic()

pisaster_diseasestat_calvert

#plotting Pisaster across wasting categories by season and region

season_order <- c("Winter", "Spring", "Summer", "Fall")

sample_data(pisaster_quadra_norm)$Season <- factor(sample_data(pisaster_quadra_norm)$Season, levels = season_order)

sample_data(pisaster_calvert_norm)$Season <- factor(sample_data(pisaster_calvert_norm)$Season, levels = season_order)


#plotting by season within Pisaster on Quadra
pisaster_wastingcat_quadra <-plot_ordination(
  physeq = pisaster_quadra_norm,
  ordination = diseased_pisaster_quadra_pcoa,
  color = "WastingCategory") +
 theme_classic() +  scale_color_manual(values = c("aquamarine2", "orchid1", "mediumorchid4"), name = "Wasting Category") +
  #facet_grid(SiteType~Season, axes = "all_x") +
  theme(strip.background = element_blank()) +
  theme(strip.text = element_text(size = 10)) + ggtitle("Quadra")
  

pisaster_wastingcat_quadra 


#plotting by season within Pisaster on Calvert
pisaster_wastingcat_calvert <-plot_ordination(
  physeq = pisaster_calvert_norm,
  ordination = diseased_pisaster_calvert_pcoa,
  color = "WastingCategory") +
 theme_classic() +
scale_color_manual(values = c("aquamarine2", "orchid1", "mediumorchid4"), name = "Wasting Category") +
  #facet_grid(SiteType~Season, axes = "all_x")  +
  theme(strip.background = element_blank()) +
  theme(strip.text = element_text(size = 10)) + ggtitle("Calvert")
  
pisaster_wastingcat_calvert

wastingcat_pcoa_region<-ggarrange(pisaster_wastingcat_calvert, pisaster_wastingcat_quadra, ncol=2, align = "v", legend = "bottom", common.legend=TRUE)

wastingcat_pcoa_region

#save plot to combine with others in another script
saveRDS(wastingcat_pcoa_region, file = "C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/twastingcat_pcoa_region.rds")

# Plot of Pisaster based on swab type  within each region
pisaster_swabtype_quadra <-plot_ordination(
  physeq = pisaster_quadra_norm,
  ordination = diseased_pisaster_quadra_pcoa,
  color = "DermisType2") +
 scale_color_manual(values = c("olivedrab", "deepskyblue", "deepskyblue4"), name = "Tissue Type", labels = c("AsympH", "WastingNL", "WastingL")) + 
  theme_classic() +
   #facet_grid(SiteType~Season, axes = "all_x") +
  theme(strip.background = element_blank()) +
  theme(strip.text = element_text(size = 10)) +
  ggtitle("Quadra")

pisaster_swabtype_quadra

pisaster_swabtype_calvert <-plot_ordination(
  physeq = pisaster_calvert_norm,
  ordination = diseased_pisaster_calvert_pcoa,
  color = "DermisType2") +
 scale_color_manual(values = c("olivedrab", "deepskyblue", "deepskyblue4"), name = "Tissue Type", labels = c("AsympH", "WastingNL", "WastingL")) + 
  theme_classic() +
  # facet_grid(SiteType~Season, axes = "all_x") +
  theme(strip.background = element_blank()) +
  theme(strip.text = element_text(size = 10)) + geom_point(size = 0.5) +
  ggtitle("Calvert")

pisaster_swabtype_calvert

tissuetype_pcoa_region<-ggarrange(pisaster_swabtype_calvert, pisaster_swabtype_quadra, ncol=2, align = "v", legend = "bottom", common.legend=TRUE)

tissuetype_pcoa_region

#save plot to combine with others in another script
saveRDS(tissuetype_pcoa_region, file = "C:/Users/Carolyn/Documents/seastar-microbiome-paper/code/tissuetype_pcoa_region.rds")


#trying another ordination with just H_diseased and L_diseased
#here we want to just compare H_diseased and L_diseased so ignore H_healthy, separately within each region

pisaster_diseased_quadra <- ps_filter(decontam_filter_nonegs_prev,
  (Host == "Pisaster ochraceus" & DermisType2 != "H_healthy" & SiteType == "Quadra"))

#check if it worked
table(sample_data(pisaster_diseased_quadra)$DermisType2)
  
pisaster_disease_norm_quadra <- pisaster_diseased_quadra %>%
  transform_sample_counts(function(x) {x/sum(x)})

#going with bray and PCOA
diseased_only_pcoa_quadra<- ordinate(pisaster_disease_norm_quadra, method = "PCoA", distance = "bray")


pisaster_diseaseonly_quadra <-plot_ordination(
  physeq = pisaster_disease_norm_quadra,
  ordination = diseased_only_pcoa_quadra,
  color = "DermisType2", 
  shape = "WastingCategory") +
 scale_color_manual(values = c("lightgreen", "deepskyblue4")) + 
   theme_classic2() +
  theme(
    strip.background = element_blank() # Removes grey fil   # Removes border
  ) + facet_grid(Season~Year, axes = "all_x")

pisaster_diseaseonly_quadra

#could do same for calvert, but leaving here as probably won't include this

```

Now some stats for all three categories examined above

```{r}
#using adonis to run a PERMANOVA by wasting cat, separately by region

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_wasting_c <- phyloseq::distance(pisaster_calvert_norm, method = "bray")

# make a data frame from the sample_data
pis_df_wasting_c <- data.frame(sample_data(pisaster_calvert_norm))


# Permanova by WASTING CAT, Quadra
adonis2(dist_mat_wasting_c~ WastingCategory, data = pis_df_wasting_c)

bd_wastingcat_calvert <- betadisper(dist_mat_wasting_c, pis_df_wasting_c$WastingCategory)
anova(bd_wastingcat_calvert) #p = 0.05052


plot(bd_wastingcat_calvert)



####### and by SWAB TYPE, Quadra
pis_df_wasting_c

adonis2(dist_mat_wasting_c~ DermisType2, data = pis_df_wasting_c)

#beta dispersion test
bd_tissuetype_c<- betadisper(dist_mat_wasting_c, pis_df_wasting_c$DermisType2)
anova(bd_tissuetype_c) # P =  0.009

plot(bd_tissuetype_c)


```


Same thing but for Calvert

```{r}
#using adonis to run a PERMANOVA by wasting cat, separately by region

set.seed(1)

# Calculate bray curtis distance matrix
dist_mat_wasting_q <- phyloseq::distance(pisaster_quadra_norm, method = "bray")

# make a data frame from the sample_data
pis_df_wasting_q <- data.frame(sample_data(pisaster_quadra_norm))


# Permanova by WASTING CAT, Quadra
adonis2(dist_mat_wasting_q~ WastingCategory, data = pis_df_wasting_q)

bd_wastingcat_quadra <- betadisper(dist_mat_wasting_q, pis_df_wasting_q$WastingCategory)
anova(bd_wastingcat_quadra) #p = 0.003

TukeyHSD(bd_wastingcat_quadra)

plot(bd_wastingcat_quadra)



####### and by SWAB TYPE, Quadra
pis_df_wasting_q

adonis2(dist_mat_wasting_q~ DermisType2, data = pis_df_wasting_q)

#beta dispersion test
bd_tissuetype_q<- betadisper(dist_mat_wasting_q, pis_df_wasting_q$DermisType2)
anova(bd_tissuetype_q) # P =  0.009

plot(bd_tissuetype_q)

```
